Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-02-08 man/ms.Rmd
change

The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. 2005 and 2012 were considered as single drought events. The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Appendix S6) and this length was used in other studies [*e.g.* @Gazol2017]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 and compared them with drought severity (Figure S5).

@@

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012; @Wilcox2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` [@Mair2017] and `rcompanion` [@Mangiafico2017] R packages. The level of significance was set to 0.05 and adjusted for multiple comparisons. All analyses were carried out in R software [@R2017] using several packages (see Supplementary `REVIEW` S?)

@@

Our findings show that severe droughts, such us 2005 and 2012, provoked a reduction both in greenness (*i.e.* primary growth) and in secondary growth of *Q. pyrenaica* (Figures 2 and 4). These results are consistent with previous studies that reported reductions of tree-growth of this species during extreme drought events, particularly during 1995 drought [@Corcuera2006; @GeaIzquierdo2014; @RubioCuadrado2018]. At this pointer-year, we observed the greatest reduction of tree-growth in our study sites, in concordance with findings reported for several Mediterranean-tree species [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018].

Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed positive trend  during the 2000 to 2016 period). This confirms previous findings which suggests an increase in primary productivity for this forests [@PerezLuque2015onto], considered the most productive ecosystems in this mountain region [@Alcaraz2016obsnev_ndvi]. 
For tree-growth, positive trends were also observed in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014]. This pattern contrasts with the decline trends observed for other mediterranean [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015] and temperate [*e.g*, *Fagus sylvatica*; @Dorado2017] tree-species located in their rear-edges.

Resilience (*Rs*) values observed in our study sites for the latest drought events in both tree-growth and greenness were close to 1, except for the tree-growth during 2005 event (Tables S1 and S2). Despite the 2012 drought event was more severe and intense than 2005, according with our analysis with SPEI data (Table S3), resilience values for greenness and tree-growth were greater for 2012 than for 2005. These results could be explained by the different seasonality of the two droughts, *i.e.*, the 2012 drought was a winter drought [@Trigo2013] that occurred earlier than the 2005 drought. The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q. pyrenaica* forests [@PeresLis2017]. Algunos autores han apuntado la importancia de los legacies (i.e. cuanto tiempo tardan en recuperarse los individuos tras una sequía) (Anderrbegg… CITAS). Nuestros resultados sugieren que otro aspecto adicional a considerar al analizar los impactos de las sequías sobre las especies de árboles es el momento en el que ocurre la sequía.

>>>>

The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. 2005 and 2012 were considered as single drought events. The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 and compared them with drought severity (Figure S5).

@@

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012; @Wilcox2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` [@Mair2017] and `rcompanion` [@Mangiafico2017] R packages. The level of significance was set to 0.05 and adjusted for multiple comparisons. All analyses were carried out in R software [@R2017] using several packages (see Appendix S7).

@@

Our findings show that severe droughts, such us 2005 and 2012, provoked a reduction both in greenness (*i.e.* primary growth) and in secondary growth of *Q. pyrenaica* (Figures 2 and 4). These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014; @RubioCuadrado2018]. The greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across Iberian Peninsula [@Penuelas2001; @Camarero2018; @Gazol2018].

During the last decades, the south of Iberian Peninsula suffered two of the worst recorded drought [2005 and 2012; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3). However, we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed positive trend  during the 2000 to 2016 period). This confirms previous findings findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014]. Our pattern differs from the decline trends observed for several mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

Resilience (*Rs*) values observed in our study sites for the latest drought events in both tree-growth and greenness were close to 1, except for the tree-growth during 2005 event (Tables S1 and S2). Despite the 2012 drought event was more severe and intense than 2005, according with our analysis with SPEI data (Table S3), resilience values for greenness and tree-growth were greater for 2012 than for 2005. These results could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] that occurred earlier than the 2005 drought. The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q. pyrenaica* forests [@PeresLis2017]. Our results highlight the importance of the drought timing on the tree-growth responses, as have been indicated for other Mediterranean-tree species [*e.g.* *Q. ilex*, @Camarero2015b].

As we mentioned above, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3). This pattern was also observed for several *Pinus* species in their rear edge [@SanchezSalguero2012; @Camarero2015]. But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6). In addition, the analysis from a long-term perspective of the severe droughts suggests a positive relation between the tree-growth resilience and drought severity (Figures 6). Surprisingly this result contradicts our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge.

Several considerations have to be taken into account to explain this result. First of all we need to explore the cumulative precipitation of the previous and posterior years of these drought events. For instance, the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern. In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009]. In addition, this argument could also explain the higher values of recovery found for 1995 (Figure 6).
